---
output: github_document
bibliography: inst/references.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# PredictionMatching
<!-- badges: start -->
<!-- badges: end -->


**PredictionMatching** provides weighted evaluation for cross-validation using raking when training data are sampled non-representatively.

> *Experimental*: This is an implementation of weighted evaluation based on raking (@Brenning2026).

## Installation

You can install the development version of PredictionMatching from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("JanLinnenbrink/PredictionMatching")
```

## Example

### Data and setup

We use an example where the sampling effort is biased along a predictor included in the model (e.g., towards low elevation) to demonstrate the weighting functions implemented in this package:

```{r example}
#| message: false
library(PredictionMatching)
library(terra)
library(sf)
library(simsam)
library(caret)
library(cowplot)

set.seed(100)

calc_rmse <- function(pred, obs) {
  sqrt(mean((pred - obs)^2, na.rm = TRUE))
}

raster_stack <- terra::rast(system.file(
  "extdata",
  "rasters_example.tif",
  package = "PredictionMatching"
))
data(train_data)

predictor_names <- setdiff(names(raster_stack), "outcome")
predictor_stack <- raster_stack[[predictor_names]]

terra::plot(predictor_stack[["elev"]])
terra::plot(vect(train_data), add = T)
```


### kNNDM cross-validation

We train a random forest model and assess its performance using kNNDM cross-validation (@Linnenbrink2024).

```{r}
# Split the data into folds using kNNDM
knndm_folds <- CAST::knndm(
  tpoints = train_data,
  modeldomain = predictor_stack[["elev"]],
  k = 5
)

# Train a random forest model
ctrl_knndm <- trainControl(
  method = "cv",
  index = knndm_folds$indx_train,
  indexOut = knndm_folds$indx_test,
  savePredictions = "final",
  verboseIter = FALSE
)

rf_knndm <- train(
  x = st_drop_geometry(train_data)[, predictor_names],
  y = st_drop_geometry(train_data)[["outcome"]],
  method = "ranger",
  trControl = ctrl_knndm,
  metric = "RMSE",
  num.trees = 300,
  importance = "impurity"
)

knndm_rmse <- calc_rmse(rf_knndm$pred$pred, rf_knndm$pred$obs)

plot(knndm_folds)
```

### True RMSE and RMSE from geographic kNNDM

We can also calculate the true RMSE in this setting:

```{r}
rf_pred <- predict(predictor_stack, model = rf_knndm, na.rm = TRUE)
rf_pred_vals <- terra::values(rf_pred, mat = FALSE)
outcome_vals <- terra::values(raster_stack$outcome, mat = FALSE)
true_rmse <- calc_rmse(rf_pred_vals, outcome_vals)
true_rmse
knndm_rmse
```

As can be seen by comparing the RMSE estimate obtained by kNNDM to the true RMSE, using kNNDM alone is sometimes not sufficient with biased sampling. 
This happens because kNNDM was applied in geographic space and does not necessarily correct the sampling bias along elevation.
To correct for this, we can use raking to obtain weights that correct for the sampling bias along elevation.

This process requires three steps: (1) calculate weights via `tw_calculate_weights()`, (2) compute pointwise cross-validation errors via `tw_pointwise_error()`, and (3) compute the weighted RMSE via `tw_weighted_error_stats()`.
First, we calculate weights via raking using `tw_calculate_weights()`. This function takes the training data and the predictor stack as input and returns a vector of weights that correct for the sampling bias along the specified predictors.
After obtaining weights via `tw_calculate_weights()`, we compute pointwise cross-validation errors with `tw_pointwise_error()`.
This preserves the original row IDs so that errors can be aligned correctly with weights even when caret reorders resampling rows.
Finally, `tw_weighted_error_stats()` computes the weighted RMSE.

```{r, fig.height=5, fig.width=12}
# Calculate weights from raking
w <- tw_calculate_weights(
  tpoints = st_drop_geometry(train_data)[, predictor_names],
  modeldomain = predictor_stack
)

# Standardize the pointwise errors obtained by kNNDM CV and add an ID column
pe <- tw_pointwise_error(
  obs = rf_knndm$pred$obs,
  pred = rf_knndm$pred$pred,
  id = rf_knndm$pred$rowIndex
)

# Weigh the pointwise errors using the weights obtained from raking
weighted_rmse <- tw_weighted_error_stats(w, pe)[["rmse"]]

plot_grid(
  plotlist = plot(w, pointwise_error = pe),
  nrow = 1,
  align = "v",
  labels = c("A", "B")
)
```

From panel B of the above plot we can already see that points with a higher CV-error receive a higher weight, resulting in a higher RMSE which gets closer to the true RMSE:

```{r pressure, echo = FALSE}
knitr::kable(data.frame(
  RMSE = c(true_rmse, knndm_rmse, weighted_rmse),
  row.names = c(
    "True RMSE",
    "kNNDM RMSE",
    "Weighted RMSE"
  )
))
```

## References
